The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-30 paper.tex
Conclusao - backend => back end - schizophrenic statements

% Title portion
\title{The Design, Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis resulting in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

\CEU is another Esterel-based language targeting embedded systems with novel 
functionalities:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution.
%
\item First-class synchronized timers.
\end{itemize}

In this work, we discuss the design of \CEU and propose a formal semantics of a 
small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes, which use the \emph{ATmega328} 8-bit 
microcontroller with 32 Kbytes of FLASH and 2 Kbytes of SRAM.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
ATmega328: \url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
a \C-compiled version occupies more than 2 Kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming 
languages~\cite{ceu.sensys13}.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} to an internal event is only 
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements})~\cite{esterel.schizo}.
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // awakes 1s trail and restarts the loop
        await f;
    end
end
\end{lstlisting}

Even though both sides of the \code{par/or} have an \code{await} statement to 
avoid instantaneous termination, the \code{loop} would restart instantaneously 
if the \code{await} could awake in the same reaction that reaches it.

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or making transformations in the code to 
execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop makes the periodic 
code to also execute immediately (and then in reactions to each \code{emit} 
request).
If the periodic \code{emit} depends on a condition, the code becomes more 
intricate because we need to copy the condition test to the periodic code, as 
illustrated in the right of Figure~\ref{lst.delay}.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
event void e;
par do
    loop do
        <...>
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
event void e;
var bool should_execute = false;
par do
    loop do
        <...>
        if <...> then
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...> // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Terra is a system for programming wireless sensor network applications that 
uses \CEU as its scripting language~\cite{ceu.terra}.
%
Figure \ref{fig.terra} shows the three basic elements of Terra:
\CEU as the scripting language,
the set of customized pre-built components,
and the embedded virtual-machine engine which can disseminate and install 
bytecode images dynamically.
%
This approach aims to combine the flexibility of remotely uploading code with 
the expressiveness and safety guarantees of \CEU.

@@

\CEU scripts cannot execute arbitrary \C code, instead, they rely on pre-built 
components that can be customized for different application domains.
%
Considering the domain of WSNs, Terra already provides components organized in 
four areas: radio communication, group management, data aggregation, and local 
operations (e.g., access to sensors and actuators).
%
When creating an instance of the \VM, the programmer can choose whether or not 
to include each component, setting different abstraction boundaries for 
scripts.
%
The generated \VM has to be preloaded into the embedded devices before they are 
physically distributed.

@@

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C backend, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
\C.

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provides a limited form of coroutines that supports reactive statements (e.g., 
\code{await} and \code{par/or}).

The event-driven approach~\cite{sync_async.whynotthreads} is well known and 
popular in many software communities,
such as web frameworks in the client side (e.g., \emph{jQuery}~\cite{js.jquery} 
and \emph{Rx.js}~\cite{js.rx}) and server side (e.g., 
\emph{Node.js}~\cite{js.node} and \emph{SEDA}~\cite{sync_async.seda}),
GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java 
Swing}~\cite{java.swing}),
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.

@@

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation with respect to trail allocation and 
scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \C suitable for static 
scheduling.

\subsection{Virtual Machines}

@@

\section{Final Remarks}
\label{sec.conclusion}

>>>>

% Title portion
\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis resulting in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

\CEU is another Esterel-based language targeting embedded and (soft) real-time 
systems with novel functionalities:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}

In this work, we discuss the design of \CEU and propose a formal semantics for 
a small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
Both use the \emph{ATmega328} with 32 Kbytes of FLASH and 2 Kbytes of SRAM:
\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM back end, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, a \C-compiled version occupies more than 2 Kbytes (as it includes 
all \emph{I/O} functionality preloaded with the \VM)~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, when \code{A} occurs, \code{B} is emitted (lines 
5--6) and both events become active, resulting in the invocation of \code{f()} 
and \code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} to an internal event is only 
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // without the restriction, it awakes 1s trail and restarts the loop
        await f;
    end
end
\end{lstlisting}

Even though both sides of the \code{par/or} have an \code{await} statement to 
avoid instantaneous termination, if the \code{await} could awake in the same 
reaction that reaches it, the \code{loop} would restart instantaneously 
resulting in infinite execution.

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (line 10) makes 
the periodic code to also execute immediately (line 9), and then in reactions 
to each \code{emit} request (line 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (line 2) and to copy the condition test to the periodic code 
(line 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with shizophrenic statements.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
var bool should_execute = false;
par do
    loop do
        <...>
        if <...> then
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...> // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Terra is a system for programming wireless sensor network applications that 
uses \CEU as its scripting language~\cite{ceu.terra}.
%
Figure \ref{fig.terra} shows the three basic elements of Terra:
\CEU as the scripting language,
a set of customized pre-built components,
and the embedded virtual-machine engine which can disseminate and install 
bytecode images dynamically.
%
This approach aims to combine the flexibility of remotely uploading code with 
the expressiveness and safety guarantees of \CEU.

@@

In Terra, \CEU scripts cannot execute arbitrary \C code, instead, they rely on 
pre-built components that can be customized for different application domains.
%
Considering the domain of sensor networks, Terra already provides components 
organized in four areas: radio communication, group management, data 
aggregation, and local operations (e.g., access to sensors and actuators).
%
When creating an instance of the \VM, the programmer can choose whether or not 
to include each component, setting different abstraction boundaries for 
scripts.
%
The generated \VM has to be preloaded into the embedded devices before they are 
physically distributed.

@@

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C back end, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
\C.

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model and support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provide a limited form of coroutines supporting reactive statements (e.g., 
\code{await} and \code{par/or}).

The event-driven approach of \CEU is well known~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as client and server-side web 
frameworks (e.g., \emph{jQuery}~\cite{js.jquery} and 
\emph{Node.js}~\cite{js.node}), GUI toolkits (e.g., \emph{Tcl/Tk}~\cite{tcl.tk} 
and \emph{Java Swing}~\cite{java.swing}), and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.

@@

Esterel has different compilation back ends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation with respect to trail allocation and 
scheduling:
the compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \C suitable for static 
scheduling.

@@

\section{Conclusion}
\label{sec.conclusion}

We presented the design, semantics, and implementation of \CEU, a synchronous 
reactive language based on Esterel targeting constrained embedded systems.

\CEU is a concurrency-safe language, employing a static analysis that encompass 
all control constructs and ensures that the high degree of concurrency in 
embedded systems does not pose safety threats to applications.
%
As a summary, the following safety properties hold for all programs that 
successfully compile in \CEU:
time and memory-bounded reactions to the environment (except for external 
system calls),
no race conditions in shared memory,
reliable abortion for activities handling resources,
and automatic synchronization for timers.
%
These properties are usually desirable in applications and are guaranteed as 
preconditions in \CEU by design.

\CEU is a resource-efficient language suitable for constrained embedded 
systems.
The reference implementation compiles to portable event-driven code in \C, with 
no special requirements for OS threads or per-trail data stacks.
The \VM implementation uses the same front end and imposes no extra 
restrictions, being equally suitable for constrained systems.

\CEU is a practical language with expressive control constructs, such as 
lexically scoped parallel compositions, convenient first-class timers, and a 
unique stack-based signaling mechanism.
%
Programs interoperate seamlessly with \C, and can take advantage of existing 
libraries, lowering the entry barrier for adoption.
%
\CEU has an open source implementation and bindings for \emph{TinyOS}, 
\emph{Arduino}, and the \emph{SDL} graphical library.%
\footnote{Website of \CEU: \url{http://www.ceu-lang.org/}}

For the past three years, we have been teaching \CEU for undergraduate and 
graduate students in research projects and two hands-on courses on 
\emph{distributed systems} and \emph{reactive programming}.
%, in which the studnets designing protocols, embedded systems, and graphical 
%applications.
%
Our experience shows that students take advantage of the sequential-imperative 
style of \CEU and can implement non-trivial concurrent applications in a few of 
weeks.

% Idexx

\begin{comment}
We are also extending and applying \CEU in more dynamic domains, such as GUIs 
and games~\cite{ceu.mod15}, by relaxing some restrictions in the language.
In particular, programs in these domains typically require unbounded memory and 
execution time, such as for dealing with dynamic collections.
We provide unsafe annotations to circumvent the rigid semantics of \CEU, 
requiring them to be explicit and trackable at the same time.
\end{comment}
